---
title: "machine learning project"
author: "Fan Ouyang"
date: "April 10, 2017"
output:
  html_document: default
  pdf_document: default
---

###Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Introduction
### Variables
Outcome variable is classe, a factor variable with 5 levels. For this data set, “participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions: 
- exactly according to the specification (Class A)
- throwing the elbows to the front (Class B)
- lifting the dumbbell only halfway (Class C)
- lowering the dumbbell only halfway (Class D)
- throwing the hips to the front (Class E)

### Models
Two models will be tested using decision tree and random forest algorithms. The model with the highest accuracy will be chosen as our final model.

## load data and cleaning data 
```{r}
library(caret)
trnLnk <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
tstLnk <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training_data <- read.csv(url(trnLnk),na.strings=c("NA","#DIV/0!",""),header=T)
testing_data <- read.csv(url(tstLnk),na.strings=c("NA","#DIV/0!",""),header=T)

#Check dimensions of train dataset
dim(training_data)
dim(testing_data)

##Create list of unwanted fields:
trnRemCols <- grepl("^X|timestamp|window|user_name",names(training_data))
tstRemCols <- grepl("^X|timestamp|window|user_name",names(testing_data))

#Remove unwanted fields
trnRmUnwtdCols <- training_data[,!trnRemCols]
tstRmUnwtdCols <- testing_data[,!tstRemCols]
#Create list of near zero variance fields
NearZeroVar <- nearZeroVar(trnRmUnwtdCols,saveMetrics=T)

#Remove near zero variance fields
trnRmZVCols <- trnRmUnwtdCols[,!NearZeroVar$nzv]
tstRmZVCols <- tstRmUnwtdCols[,!NearZeroVar$nzv]

#Remove fields with NAs
trnNArmCondn <- (colSums(is.na(trnRmZVCols))!=0)
tstNArmCondn <- (colSums(is.na(tstRmZVCols))!=0)

trnRmNACols <- trnRmZVCols[,!trnNArmCondn]
tstRmNACols <- tstRmZVCols[,!tstNArmCondn]

#New Training and Testing Datasets after clean-up
trnDataNew <- trnRmNACols
tstDataNew <- tstRmNACols

dim(trnDataNew); dim(tstDataNew)
```
## prediction
The datasets are now partitioned to make use of cross validation
```{r}
set.seed(1234) 

#Train-model Validation Partition
intrain <- createDataPartition(y=trnDataNew$classe,p=0.7,list=F)

modTRNSample <- trnDataNew[intrain,] #To be used for model training
modTSTSample <- trnDataNew[-intrain,] #To be used for testing accuracy of models

dim(modTRNSample); dim(modTSTSample)
```
### building models
## Method 1: Decision Tree
```{r}
library(rpart)
model1 <- rpart(classe~.,method="class",data=modTRNSample)
prediction1<-predict(model1,modTSTSample,type="class")
library(rpart.plot)
rpart.plot(model1,main="classification tree",extra=102,under=T,faclen=0)
confusionMatrix(prediction1,modTSTSample$classe)
```
## Method 2: Random Forest
```{r}
library(randomForest)
model2 <- randomForest(classe~.,method="class",data=modTRNSample)
prediction2<-predict(model2,modTSTSample,type="class")
confusionMatrix(prediction2,modTSTSample$classe)
```

### Conclusion
Random forest, though a little more complex, was way more accurate. Hence, the random forest technique was chosen as the final prediction algorithm.

### submission
```{r}
predictfinal <- predict(model2, testing_data, type="class")
predictfinal
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)
```
